A Tutorial on Tailored Simulation-Based Power Analysis for Experimental Designs with Generalized Linear Mixed Models

### 2024-04-10 docs/manuscript.Rmd
delete the old case study chapter that has now been fully integrated into the new structure

# THIS WAS THE OLD CASE STUDY CHAPTER

<!-- overview of this section -->

In this section, we outline the steps for performing data simulation and a priori power analysis for GLMMs using a case study based on a specific experimental study design from the area of human-AI (artificial intelligence) interaction research. All code in this manuscript and simulation results are available in the project's repository on the Open Science Framework (https://osf.io/dhwf4/).

## Experimental study design

<!-- describe the study idea and design -->
In the present case study, we simulate data for an experiment where the diagnostic performance of users of an AI-enabled diagnostic decision support system will be evaluated. 
The goal is to understand how AI advice influences medical decision-making. Participants, radiologists (task experts) and students/interns (non-task experts), review head computer tomography (CT) scans to assess the presence of a bleeding. 
To support their decision-making, an AI model provides initial diagnostic advice, which can be used as guidance by the participants. This AI advice can be either correct (80% of cases) or incorrect (20%). In the control condition, no AI advice is presented, meaning that the participants have to read the CT scan without any support. After reviewing the CT scan, participants deliver a medical diagnosis (bleeding or no bleeding), which may be either accurate or inaccurate.
This experimental design introduces some missing values by design since the advice is neither correct nor incorrect when no advice is present, which must be taken into account when simulating and analyzing the data. With this experiment, we want to determine if (a) experts are better than non-experts in reading head CT scans and if (b) correct AI advice leads to better diagnostic accuracy than incorrect AI advice.
In this example, recruiting task experts (i.e., radiologists) is more challenging due to their limited availability, while non-experts (i.e., students/interns) are more readily accessible. 
The goal of the present simulation-based power analysis is to determine how many task experts and non-experts must be recruited to achieve sufficient statistical power in the planned experiment.

## Our specific GLMM

<!-- deep dive into our specific example -->
In a GLMM, the expected value of the dependent variable Y conditioned on the vector of predictor variables $\mathbf{X}$ and random effects $\mathbf{U}$, transformed by a link function $g()$ is modeled as a linear combination $\eta$ of the predictor variables $\mathbf{X}$, the random effects $\mathbf{U}$ and the model parameters $\mathbf{\beta}$ [@fahrmeirRegressionModelsMethods2021]: 
$$
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta
$$
Equivalently, the conditional expected value is modeled as the linear combination $\eta$, transformed by the inverse link function $g^{-1}()$: 
$$
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u}) = g^{-1}(\eta)
$$
If the dependent variable (i.e., diagnostic decision) $Y$ is a binary variable with values $0$ (i.e., inaccurate), or $1$ (i.e., accurate), the conditional expected value is equivalent to the probability: 
$$
P_{si} := P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
$$
In our case study, $P_{si}$ is the conditional probability that a subject $s$ gives the correct response to item (i.e., CT scan) $i$.

In such a setting, we model this probability as
$$
P_{si} = inverse\_logit(\eta_{si})
$$
with the inverse-logit link $g^{-1}(\eta_{si}) = inverse\_logit(\eta_{si}) = \frac{exp(\eta_{si})}{1 + exp(\eta_{si})}$ or equivalently
$$
logit(P_{si}) = \eta_{si}
$$
with the logit link $g(P_{si}) = logit(P_{si}) = ln (\frac{P_{si}}{1 - P_{si}})$.

In total, we assume the model
$$
\begin{aligned}
logit[P_{si}] =\ (&\beta_0 + u_{0s} + u_{0i}) + \\
&\beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}
\end{aligned}
$$
or equivalently
$$
\begin{aligned}
P_{si} = inverse\_logit[&(\beta_0 + u_{0s} + u_{0i}) + \\
&\beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$
with model parameters $\beta_0$, $\beta_e$, $\beta_a$, $\beta_c$, $\beta_{ea}$, $\beta_{ec}$, $\sigma_S$, and $\sigma_I$.

## Data simulation

In the first six lines of the function definition, we set some default parameter values (which we will explain in a later section) and load the packages we use to manipulate and simulate data. 
In our case study, each subject (`n_subjects` in total) is assumed to respond to each item (i.e., CT scan; `n_items` in total). Thus, the `add_random` command creates a fully-crossed `data.frame` with `n_subjects` $\times$ `n_items` rows. We add a between-subject effect with the `add_between` command, simulating that about $25\%$ of subjects are experts. The next two lines simulate that in $\frac{2}{3}$ of trials, subjects will be presented with AI advice, and if advice is presented, the advice will be correct in about $80\%$ of cases (the variable `advice_correct` is always 0 when no advice is presented). Next, we simulate one random effect for each subject (`u0s`) and for each item (`u0i`). As assumed by standard GLMMs, the `add_ranef` function draws the random effects from a normal distribution with a mean 0 and a standard deviation specified by the user. With all design variables done, we are ready to simulate our model equation outlined in the last section. The linear predictor variable `linpred` ($\eta$ in the GLMM model equations) combines the predictor variables, random effects, and model parameters as assumed by our model. We then transform the linear predictor with the inverse-link function to compute `y_prob`, the probability that the subject correctly solved the item (in R, the inverse-logit link is computed with `plogis` and the logit link with `qlogis`). In the final step, we simulate the binary dependent variable `y_bin` (i.e., whether the subject makes an accurate diagnostic decision for the CT scan) by -- for each trial -- drawing from a Bernoulli distribution with success probability `y_prob`.

## Model fitting

<!-- ## The lme4 package in R -->

<!-- describe the lme4 package and how it works-->
In our case study, we use the lme4 R package [@batesFittingLinearMixedEffects2015], which is a state-of-the-art tool for fitting frequentist GLMMs.[^1]
The lme4 package includes a function called `simulate` that allows researchers to simulate the dependent variable based on the same model formula used for model fitting, enabling simulation-based power analyses and other related analyses.

However, the model parameterization used by the lme4 package is quite technical, making it difficult for applied researchers to determine whether their specified population model (i.e., the theoretical model that describes the underlying data generation process for a specific population of interest) implies plausible associations in their simulated data. Therefore, in this tutorial, we simulate data for GLMMs from first principles (i.e., creating synthetic data step by step instead of using black box functions) to assist applied researchers in better understanding all model assumptions and then use lme4 to analyze the simulated datasets.[^2]

In this section, we show how to fit a GLMM with lme4, interpret the model, and test hypotheses derived from a research question.
We simulate data according to our model, in which 100 subjects respond to 50 items (we use `set.seed` to make the simulation reproducible). However, for the sake of the exercise, we can imagine that this would be real data resulting from our future experiment and think about how we would analyze this data.

## Model interpretation

In the output, the `Estimate` column in the `Fixed effects` table contains the estimates for the $\beta$ parameters, while the `Std.Dev.` column in the `Random effects` table contains the estimates for $\sigma_S$ and $\sigma_I$.

Unfortunately, the model parameters in a binomial GLMM are hard to interpret because 1) the $\beta$ parameters are connected to the modeled probability via the non-linear inverse-logit link, and 2) we also have to consider the random effects. The most simple interpretation works by imagining a subject with average ability ($u_{0s} = 0$) responding to an item (i.e., CT scan) with average difficulty ($u_{0i} = 0$). Then the model implied probability that such a person solves such an item accurately is given by: 
$$
\begin{aligned}
P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0}) = \\
= inverse\_logit[&\beta_0 + \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$

In fact, we would only need the full equation if the subject is an expert and correct advice is presented. In all other experimental conditions, some terms drop from the equation because they are multiplied by $0$. The other extreme case would be the probability that a non-expert with average ability solves an item with average difficulty when no advice is presented:
$$
\begin{aligned}
P(Y=1|expert = 0, advice\_present = 0, advice\_correct = 0, u_{0s} = 0, u_{0i} = 0) = \\
= inverse\_logit[\beta_0]
\end{aligned}
$$

Due to this complicated relationship, we argue not to focus too much on interpreting single model parameters when working with GLMMs. Instead, it can be more intuitive to consider model predictions and the model-implied distribution of the dependent variable for each experimental condition across all subjects and items.

With the marginaleffects package [@R-marginaleffects], we can easily compute predictions for all observations in the dataset based on the fitted GLMM (including all fixed **and** random effects), and plot the average probability with confidence intervals for each experimental condition in Figure \@ref(fig:marginaleffects):

(ref:marginaleffects) Marginal distributions including means and 95% confidence intervals for all experimental conditions computed with the marginaleffects package.

```{r marginaleffects, message=FALSE, fig.cap="(ref:marginaleffects)"}
library(marginaleffects)
plot_predictions(fit, by = c("advice_present", "advice_correct", "expert"),
  type = "response") + ylim(c(0.3, 1))
```

## Hypothesis testing

However, we need to think about the model parameters again when we want to test hypotheses that we have theoretically derived from some research question. Because the inverse-logit link is still a continuously increasing function, positive parameter values always correspond to increases in probability and vice versa.

The `Fixed effects` table in the lme4 summary output also includes p-values for hypothesis tests with null hypotheses of the style $H_0: \beta = 0$. However, for many research questions of interest, we are not interested in these two-sided tests that refer to only a single parameter.

For our case study, imagine the following combined hypothesis: *We expect that for both experts and non-experts, correct advice leads to a higher probability of accurately diagnosing a CT scan compared to no advice presented, AND, we expect that for both experts and non-experts, incorrect advice leads to a lower probability of accurately diagnosing a CT scan compared to no advice presented.*

This combined hypothesis leads to the following four separate null hypotheses to be tested: $$
\begin{aligned}
H_{01}:\ &\beta_{a} + \beta_{c} + \beta_{ea} + \beta_{ec} \leq 0 \\
H_{02}:\ &\beta_{a} + \beta_{c} \leq 0 \\
H_{03}:\ &\beta_{a} + \beta_{ea} \geq 0 \\
H_{04}:\ &\beta_{a} \geq 0
\end{aligned}
$$

We arrive at these inequalities based on the following logic, exemplified here only for $H_{01}$: The first null hypothesis states that *an expert responding to an item while presented with correct advice has a lower or equal probability of solving the item compared to the same expert facing the same item without any advice.* This implies the following inequality for each subject $s$ and item $i$ $$
inverse\_logit[(\beta_0 + u_{0s} + u_{0i}) + \beta_e + \beta_{a} + \beta_{c} + \beta_{ea} + \beta_{ec}] \leq inverse\_logit[(\beta_0 + u_{0s} + u_{0i}) + \beta_e]
$$ which simplifies to $\beta_{a} + \beta_{c} + \beta_{ea} + \beta_{ec} \leq 0$.

We can specify and test hypotheses like these with the multcomp package [@hothornSimultaneousInferenceGeneral2008] as follows:

```{r, message=FALSE}
library(multcomp)
null_hypotheses <- c(
  "advice_present + advice_correct + expert:advice_present + 
   expert:advice_correct <= 0",
  "advice_present + advice_correct <= 0",
  "-1 * (advice_present + expert:advice_present) <= 0",
  "-1 * (advice_present) <= 0")
glht <- glht(fit, linfct = null_hypotheses)
summary(glht, test = univariate())$test$pvalues
```

Because all hypotheses tested simultaneously with the `glht` function must have the same direction, we flip the sign of inequalities three and four by multiplying them with $-1$. The multcomp package automatically adjusts p-values when multiple hypotheses are tested simultaneously [@hothornSimultaneousInferenceGeneral2008]. However, the combined null hypothesis in our exemplary research question should only be rejected if **all** individual null hypotheses are rejected [i.e., intersection-union setting; @dmitrienkoTraditionalMultiplicityAdjustment2013]. In such cases, the error probabilities do not accumulate, and we would waste power when correcting for multiple tests. Thus, we request unadjusted p-values by setting `test = univariate()` in the `summary` command. With a standard significance level of $\alpha = 0.05$, we would not reject all four null hypotheses (the p-value for hypothesis $H_{03}$ is not significant) and therefore also not reject the combined null hypothesis for this simulated dataset. Note that this decision would be wrong because we have simulated the data such that the combined alternative hypothesis is actually true in the population.

## Specification of plausible parameter values

When introducing our simulation function and simulating data for the above example, we have used theoretically plausible values as defaults for all model parameters ($\beta_0$, $\beta_e$, $\beta_a$, $\beta_c$, $\beta_{ea}$, $\beta_{ec}$, $\sigma_S$, and $\sigma_I$), but have not talked about where these numbers came from.

Ideally, one would rely on meta-analytic results or conclusive data from pilot studies. However, these are sometimes not readily available. All parameter values in our present case study have been determined based on results from related prior work. Additionally, we had repeated discussions with our affiliated domain experts in radiology to check our assumptions.

We now outline a few strategies on how to determine plausible parameter values. We have already seen in our discussion of model interpretation how we can derive the model implied probability for each experimental condition, that a subject with average ability solves an item with average difficulty. We can revert this perspective by choosing plausible probability values and deriving the parameter values implied by these probabilities (for an average subject and an average item).

```{r, echo=FALSE}
probtable <- data.frame(
  col1 = 
    c("no advice, no expert", "no advice, expert", "false advice, no expert", 
      "false advice, expert", "correct advice, no expert", "correct advice, expert"),
  col2 =
    c("0.70", "0.90", "0.40", "0.85", "0.90", "0.95"),
  col3 = 
    c("$logit(0.70) = \\beta_0$",
      "$logit(0.90) = \\beta_0 + \\beta_e$", 
      "$logit(0.40) = \\beta_0 + \\beta_a$",
      "$logit(0.85) = \\beta_0 + \\beta_e + \\beta_{a} + \\beta_{ea}$",
      "$logit(0.90) = \\beta_0 + \\beta_a + \\beta_c$",
      "$logit(0.95) = \\beta_0 + \\beta_e + \\beta_a + \\beta_c + \\beta_{ea} + \\beta_{ec}$"))
apa_table(
  probtable
  , col.names = c("Experimental condition",
    "$P(Y=1|\\mathbf{X=x}, \\mathbf{U} = \\mathbf{0})$",
    "Implied equation")
  , caption = "Assumed probabilities that an average subject solves an average item in each experimental condition."
  , note = "Implied equations are derived based on the model equations and setting all random intercept terms to 0."
  , escape = FALSE
)
```

Although the derivations above are straightforward, it is important not to misinterpret their implications: In binomial GLMMs, the average probability to solve an item (averaged across persons of varying ability and items of varying difficulty) is **not** equal to the probability that a person with average ability solves an item with average difficulty. The first perspective implies a so-called marginal interpretation, while the second one implies a conditional interpretation. For example, we determined the $\beta$ parameters in a way that corresponds to a desired conditional probability of $0.95$, that an expert with average ability solves an item with average difficulty when presented with correct advice. However, even if the model were true, we would not observe this probability value if we estimated the marginal probability in a group of experts responding to items presented with correct advice from a big sample of subjects drawn from their natural distribution of ability and items drawn from their natural distribution of difficulty.

The inequality of conditional and marginal effects in GLMMs [@fahrmeirRegressionModelsMethods2021] makes their interpretation more difficult. One must be careful when specifying parameter values based on previous studies or pilot data that use the marginal interpretation (e.g., a pilot study providing an estimate of how often neurologists make an accurate diagnosis based on brain scans). However, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values: When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge. Then, we can iteratively adjust the parameter values until we are satisfied with the implied distributions.

Earlier, we have already encountered one way to visualize the implied marginal distributions: We can fit our model to a simulated dataset and use the convenience functions from the marginaleffects package to compute averaged predictions that correspond to our quantities of interest. However, the model predictions will only be close to the true distribution if the simulated dataset is very large, but then the model fitting consumes a lot of time and memory. A more sophisticated strategy is to simulate a large dataset and directly compute the averages, contrasts and distributions we are interested in.

```{r}
library(tidyverse)
library(ggdist)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 0.5, sd_u0i = 0.5)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

```{r}
library(tidyverse)
library(ggdist)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 0.5, sd_u0i = 0.5)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>%
  group_by(condition) %>%
  summarize(freq_y_1 = sum(y_bin) / n())
```

Up to this point, we have not talked about plausible values for the standard deviations of the subject and item random intercepts ($\sigma_S$ and $\sigma_I$). Plots like the one above are a useful tool to decide whether the specified standard deviations are reasonable by comparing the ranges and overlap between conditions to domain knowledge.

```{r}
dat <- simulate(n_subjects = 2000, n_items = 2000, sd_u0i = 0.01)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

```{r, include=FALSE}
dat <- simulate(n_subjects = 2000, n_items = 2000, sd_u0s = 0.01)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

```{r}
dat <- simulate(n_subjects = 2000, n_items = 2000, sd_u0s = 3, sd_u0i = 3)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

## Results

With all these considerations addressed, we are finally ready to perform a power analysis. Wrapping the `simulate` function already constructed earlier, the helper function `sim_and_analyse` performs all previous steps (simulate a dataset, fit a GLMM, compute p-values) in a single command.

```{r}
sim_and_analyse <- function(
  formula_chr = "y_bin ~ 1 + expert + advice_present + advice_correct + 
    expert:advice_present + expert:advice_correct + (1|subject) + (1|item)", 
  null_hypotheses = c("advice_present + advice_correct + 
    expert:advice_present + expert:advice_correct <= 0",
    "advice_present + advice_correct <= 0",
    "-1 * (advice_present + expert:advice_present) <= 0",
    "-1 * (advice_present) <= 0"), ...){
  require(lme4)
  require(multcomp)
  # simulate data
  dat <- simulate(...)
  # fit model
  model <- glmer(as.formula(formula_chr), data = dat, family = "binomial")
  # compute p-values
  glht <- glht(model, linfct = null_hypotheses)
  pvalues <- summary(glht, test = univariate())$test$pvalues
  setNames(pvalues, paste0("p_H0", 1:length(null_hypotheses)))
}
```

Power analysis can quickly become computationally intensive when we repeatedly simulate data and fit models for different parameter combinations or sample sizes. Thus, we use the future [@R-RJ-2021-048] and furrr [@R-furrr] packages to perform computations in parallel. First, we enable parallelization and specify how many parallel cores ("workers") of our computer to use (users can find out the maximum number of cores on their computer with the command `parallel::detectCores()`), and set a seed to make the simulation reproducible.

```{r, message=FALSE, echo=FALSE, eval=FALSE}
library(future)
library(furrr)
plan("multisession", workers = 6)
set.seed(2)
sim_design <- crossing(
  rep = 1:300,
  n_subjects = c(100, 150, 200, 250),
  n_items = c(10, 30, 50, 70)
) %>%
  mutate(pvalues = future_pmap(., sim_and_analyse, 
    .options = furrr_options(seed = TRUE))) %>%
  unnest_wider(pvalues)
saveRDS(sim_design, file = "results.rds")
```

```{r, message=FALSE, eval=FALSE}
library(furrr)
sim_design <- crossing(
  rep = 1:300,
  n_subjects = c(100, 150, 200, 250),
  n_items = c(10, 30, 50, 70)
) %>%
  mutate(pvalues = future_pmap(., sim_and_analyse, 
    .options = furrr_options(seed = TRUE))) %>%
  unnest_wider(pvalues)
```

The result of the computation is a data frame that contains the p-values of all tested hypotheses for each simulated dataset.
In some iterations (predominantly in conditions with small sample sizes), model estimation did not converge with the lme4 package. When the model fails to converge, it means that the statistical model being fitted to the data failed to reach a stable or valid solution during the estimation process. We do not remove these results because non-convergence can also happen when analyzing the real data we plan to collect, thus, we want to factor in this possibility to keep our simulation more realistic.

For our exemplary combined hypothesis, power is defined as the (long-run) percentage of simulations in which all four p-values of our component hypotheses are significant at the $\alpha = 0.05$ level. Based on our simulation outcomes, we compute a power estimate for each combination of `n_subjects` $\times$ `n_items` (including 95% confidence intervals) and visualize the results with the following code.[^3]

```{r, echo=FALSE, eval=FALSE}
sim_design <- readRDS(file = "results.rds")
```

```{r finpwr, fig.cap="Simulation-based power estimates including 95% confidence interval of the case study for different numbers of subjects and items, based on a significance level of 0.05.", eval=FALSE}
library(binom)
alpha <- 0.05
power <- sim_design %>%
  group_by(n_subjects, n_items) %>% 
  summarise(power = mean(p_H01 < alpha & p_H02 < alpha & 
                           p_H03 < alpha & p_H04 < alpha), 
    n_sig = sum(p_H01 < alpha & p_H02 < alpha & 
                  p_H03 < alpha & p_H04 < alpha),
    n = n(),
    ci.lwr = binom.confint(n_sig, n, method = "wilson")$lower,
    ci.upr = binom.confint(n_sig, n, method = "wilson")$upper, 
    .groups = "drop")
power %>%
  mutate(across(c(n_subjects, n_items), factor)) %>%
  ggplot(aes(n_subjects, n_items, fill = power)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                power, ci.lwr, ci.upr)), 
    color = "white", size = 4) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  xlab("number of subjects") + ylab("number of items")
```

As should be the case, power estimates in Figure \@ref(fig:finpwr) increase with both the number of subjects and the number of items. The confidence intervals indicate how precisely power was estimated by our simulation. Higher precision (which would be reflected in narrower confidence intervals) could be obtained by increasing the number of repetitions (`rep`) in the simulation. In practice, data simulations are often run multiple times with adjusted combinations of sample sizes. When running for the first time, it might be revealed that power is way too low (or much higher than required) for some combinations of `n_subjects` and `n_items`. When narrowing down the best combination that achieves sufficient power while at the same time striking a good balance of how many subjects and items are practically feasible, later rounds of data simulation will typically include a smaller grid of sample sizes combined with a higher number of repetitions. This will assure high precision for the final power estimates, which are then used for the sample size justification of the future study.

## Expected effect size vs. smallest effect size of interest: sensitivity analysis

In our case study, we have performed simulation-based power analysis from a single set of parameter values that reflect our assumptions of an expected effect size. Instead of extracting this expected effect size from meta-analyses or pilot data, which has been the main focus of previous tutorials, we have demonstrated some strategies to determine plausible parameter values in GLMMs based on domain knowledge. Domain knowledge can be considered a vague theoretical model about the data-generating process that is less formal and can only be accessed by a back-and-forth exchange in which domain experts assess the plausibility of simulated data. When sample sizes are chosen based on the results of our simulation-based power analysis, a future study will be informative to reject the null hypothesis if an effect of our *expected size* is present. However, if the true effect is indeed smaller, the power will be lower, and the study might not be sufficiently informative. A common, more conservative strategy for sample size justification is to perform power analysis for the smallest effect size of interest (SESOI). An effect smaller than the SESOI would be considered too small to be interesting or practically meaningful, even if the effect is not actually zero [@kingPointMinimalImportant2011]. For strategies on the even more difficult task of specifying a plausible SESOI, as well as a thorough discussion of various topics concerning power analysis, see [@lakensImprovingYourStatistical2022]. When domain knowledge or formal theories about the research topic of interest are too vague to specify a meaningful SESOI, it is still recommended to demonstrate power for different effect sizes in what is called *sensitivity power analysis*. By simulating power for different effect sizes (in addition to the different number of subjects and items), one can make sure that power would still be sufficient to detect smaller effect sizes than our expected effect or at least get an impression of how strongly power depends on the size of the true effect. In simple study designs, it is possible to perform sensitivity power analysis based on a single standardized effect size (e.g., analyze power in a two-sample t-test for a standardized mean difference varying between 0.1 and 0.8). However, for our case study that investigates combined hypotheses in a GLMM modeling framework, the effect size is implicitly represented by the complex distribution of probabilities within and between experimental conditions. In this setting, sensitivity power analysis would require manually specifying additional sets of plausible parameter values that reflect scenarios with smaller or larger differences between groups with respect to our specific research question. Power could then be simulated for several of these scenarios (across different numbers of subjects and items, as considered earlier).

@@

1.  The growing need for simulation-based sample size planning in experimental research: In order to conduct informative research using varying complex experimental designs with GLMMs formula-based heuristics and user-friendly software tools for a priori power analysis are often not suitable. Therefore, simulation-based power analysis is becoming increasingly needed since it provides experimental researchers with a tailored approach to estimating required sample sizes before data collection.

@@

3.  Teaching data simulation skills: The ability to conduct simulation-based sample size planning is a valuable skill that should be taught to experimental researchers. By incorporating such training into research methods courses and workshops, researchers can gain a deeper understanding of statistical power and improve the quality of their experimental designs. 
Equipping researchers with the knowledge and tools to perform simulation-based sample size planning enables them to make informed decisions and enhance the rigor of their studies. 
The need to reason about how to simulate plausible data that is in line with the research hypothesis, while not violating domain expertise on how plausible data should look, might also contribute to planning more insightful studies that can answer more precise research questions [@yarkoniGeneralizabilityCrisis2022].

4.  Addressing the mismatch in effort perception: There is often a significant disconnect between the amount of effort required to perform simulation-based a priori sample size planning and the perceived effort estimated by researchers and collaborators in experimental research. Many researchers request simulation-based power analyses from statisticians or methodological experts without fully comprehending the complexity and time-consuming nature of these tailored simulations. It is crucial to raise awareness about the effort involved to ensure realistic expectations and effective collaboration between researchers and methodological experts.

5.  Recognizing the value of simulation-based design analysis: Simulation-based power analyses are not mere technicalities; they are valuable research contributions that deserve recognition in experimental research. They offer insights into the robustness and sensitivity of experimental designs, helping researchers make informed decisions about sample sizes, effect sizes, and statistical power. Their importance can be reflected by allocating them a separate publication or incorporating them as a significant component of stage 1 preregistered reports [@chambersPresentFutureRegistered2022].

@@

In the wake of the replication crisis and myriad of underpowered experimental work, generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, analytic methods and software cannot be applied to conduct a priori sample size planning for GLMMs necessitating data simulation-based approaches. Through this applied tutorial, we aim to provide researchers with the necessary skills and tools to perform simulation-based sample size planning with GLMMs themselves. By incorporating GLMMs and a priori sample size planning into their work, researchers can enhance the replicability and credibility of their experiments [@yarkoniGeneralizabilityCrisis2022].

>>>>

## Sensitivity analysis

In our case study, we have performed simulation-based sample size planning from a single set of parameter values that reflect our assumptions of an expected effect size. Instead of extracting this expected effect size from meta-analyses or pilot data, which has been the main focus of previous tutorials, we have demonstrated some strategies to determine plausible parameter values in GLMMs based on domain knowledge. Domain knowledge can be considered a vague theoretical model about the data-generating process that is less formal and can only be accessed by a back-and-forth exchange in which domain experts assess the plausibility of simulated data. When sample sizes are chosen based on the results of our simulation-based power analysis, a future study will be informative to reject the null hypothesis if an effect of our *expected size* is present (or estimate the effect with satisfying precision). However, if the true effect is indeed smaller, power (or precision) will be lower, and the study might not be sufficiently informative. A common, more conservative strategy for sample size justification is to perform sample size planning for the smallest effect size of interest (SESOI). An effect smaller than the SESOI would be considered too small to be interesting or practically meaningful, even if the effect is not actually zero [@kingPointMinimalImportant2011]. For strategies on the even more difficult task of specifying a plausible SESOI, as well as a thorough discussion of various topics concerning power analysis, see [@lakensImprovingYourStatistical2022]. When domain knowledge or formal theories about the research topic of interest are too vague to specify a meaningful SESOI, it is still recommended to demonstrate power or precision for different effect sizes in what is called *sensitivity power analysis*. By simulating power (or precision) for different effect sizes (in addition to the different number of subjects and items), one can make sure that power (or precision) would still be sufficient to detect smaller effect sizes than our expected effect or at least get an impression of how strongly power (or precision) depends on the size of the true effect. In simple study designs, it is possible to perform sensitivity analysis based on a single standardized effect size (e.g., analyze power in a two-sample t-test for a standardized mean difference varying between 0.1 and 0.8). However, for our case study that investigates combined hypotheses in a GLMM modeling framework, the effect size is implicitly represented by the complex distribution of probabilities within and between experimental conditions. In this setting, sensitivity analysis would require manually specifying additional sets of plausible parameter values that reflect scenarios with smaller or larger differences between groups with respect to our specific research question. Power (or precision) could then be simulated for several of these scenarios (across different numbers of subjects and items, as considered earlier).

@@

1.  The growing need for simulation-based sample size planning in experimental research: In order to conduct informative research using varying complex experimental designs with GLMMs, formula-based heuristics and user-friendly software tools for a priori power analysis are often not suitable. Therefore, simulation-based power analysis is becoming increasingly needed since it provides experimental researchers with a tailored approach to estimating required sample sizes before data collection.

@@

3.  Teaching data simulation skills: The ability to conduct simulation-based sample size planning is a valuable skill that should be taught to experimental researchers. By incorporating such training into research methods courses and workshops, researchers can gain a deeper understanding of statistical power or precision, and improve the quality of their experimental designs. 
Equipping researchers with the knowledge and tools to perform simulation-based sample size planning enables them to make informed decisions and enhance the rigor of their studies. 
The need to reason about how to simulate plausible data that is in line with the research hypothesis, while not violating domain expertise on how plausible data should look, might also contribute to planning more insightful studies that can answer more precise research questions [@yarkoniGeneralizabilityCrisis2022].

4.  Addressing the mismatch in effort perception: There is often a significant disconnect between the amount of effort required to perform a priori simulation-based sample size planning and the perceived effort estimated by researchers and collaborators in experimental research. Many researchers request simulation-based power analyses from statisticians or methodological experts without fully comprehending the complexity and time-consuming nature of these tailored simulations. It is crucial to raise awareness about the effort involved to ensure realistic expectations and effective collaboration between researchers and methodological experts.

5.  Recognizing the value of simulation-based design analysis: Simulation-based power analyses are not mere technicalities; they are valuable research contributions that deserve recognition in experimental research. They offer insights into the robustness and sensitivity of experimental designs, helping researchers make informed decisions about sample sizes, effect sizes, and statistical power or precision. Their importance can be reflected by allocating them a separate publication or incorporating them as a significant component of stage 1 preregistered reports [@chambersPresentFutureRegistered2022].

@@

In the wake of the replication crisis and myriad of underpowered experimental work, generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, analytic methods and software cannot be applied to conduct a priori sample size planning for GLMMs, necessitating data simulation-based approaches. Through this applied tutorial, we aim to provide researchers with the necessary skills and tools to perform simulation-based sample size planning with GLMMs themselves. By incorporating GLMMs and a priori sample size planning into their work, researchers can enhance the replicability and credibility of their experiments [@yarkoniGeneralizabilityCrisis2022].
